---
title: "Weight Lifting Exercise Quality Machine Learning Model"
author: "Jason Naramore"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

The purpose of this assignment is to quantify the quality of weight lifting exercises. Data was collected as described in the reference [1] using 3-axes acceleration, gyroscope, and magnetometer sensors attached to a subject's wrist, upper arm, waist, and dumbbell. Each subject performed Unilateral Dumbbell Bicep Curls that were classified as follows:
  
A: exactly according to the specification,  
B: throwing the elbows to the front,  
C: lifting the dumbbell only halfway,  
D: lowering the dumbbell only halfway,  
E: throwing the hips to the front.  
  
A random forest model was trained to accurately identify each of these classifications using the relevant sensor data. The out of sample  accuracy was estimated using 10% of the training data that was set aside prior to training.

## Step 1 - Load and Subset

First, the training and test sets are loaded. The training set is then subsetted to columns where there are not NA values in the testing set. Also, the first 7 columns are ignored, which contain subject identifiers, timestamps, and other unneeded variables for training. The training set is partitioned into two groups, one containing 90% of the rows to be used for training, and the remaining 10% will be used to estimate the accuracy of the model.

```{r subset}
library(caret)
training<-read.csv("pml-training.csv")
testing<-read.csv("pml-testing.csv")

#only consider columns that appear in testing
na_mask <- is.na(testing[1,])
training1<-subset(training,select = !na_mask)
training1<-subset(training1,select = c(-1:-7))

testing1 <- subset(testing,select = !na_mask)
testing1 <- subset(testing1,select = c(-1:-7))

set.seed(1234)
trainIn<- createDataPartition(training1$classe, p = .9,list=FALSE)
training2<- training1[trainIn,]
training2_test<- training1[-trainIn,]
```

## Step 2 - PCA

The 53 remaining variables in the training set are reduced to the most relevant variables using a principal components analysis. Caret's preProcess function is used, with thresh = .8 to identify the principal components that account for 80% of variability. This results in 12 components, so the model can be trained with 12 rather than 53 predictors.

```{r pca}
preProc<-preProcess(training2[,-53],method = "pca",thresh = .8)
trainPC <- predict(preProc,training2[,-53])
```

## Step 3 - Fit Random Forest Model

A random forest model is chosen to train this data, consistent with the authors reasoning in their publication [1]. Cross-validation is handled by setting the train control method to "cv". The accuracy of the final model is shown in the plot, where the resampling results indicate 96.7% accuracy with randomly selected predictors (mtry) set to 2.

```{r fit, cache=TRUE, fig.cap = "Estimated Accuracy by Resampling"}
fitControl <- trainControl(method = "cv")
set.seed(1234)
modFit<-train(y = training2$classe, x = trainPC,method = "rf",trControl = fitControl)
modFit
plot(modFit)
```

## Step 3.1 - Plot Variable Importance

The importance of each principal components is shown in the plot below.

```{r varImp, fig.cap = "Variable Importance in The Random Forest Model"}
plot(varImp(modFit))
```

## Step 4.1 - Results of Training Set

The final model classifies the training set with 100% accuracy.

```{r results_training}
train_pred <- predict(modFit,trainPC)
confusionMatrix(train_pred,factor(training2$classe))
```

## Step 4.2 - Results of Training-Test Set

Due to the high accuracy of the model, 10% of the training data was partitioned to better estimate the out of sample accuracy of the model. The accuracy of the set, called the training-test set, was 97.2%.

```{r results_training_test}
trainPCtest<-predict(preProc,training2_test[,-53])
train2test_pred <- predict(modFit,trainPCtest)
confusionMatrix(train2test_pred,factor(training2_test$classe))
```

## Step 5 - Predict Testing Set

The last step is to predict the testing set. The classe is unknown in the testing set.

```{r test_prediction}
testPC<-predict(preProc,testing1[,-53])
test_prediction <- predict(modFit,testPC)
factor_levels <- levels(factor(training$classe))
test_predict_factors <- factor_levels[test_prediction]
data.frame(problem_id = testing1$problem_id, test_predict_factors)
```

## References

[1] Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013. Website: http://groupware.les.inf.puc-rio.br/har#ixzz6kg2ypyaf
